---
title: "An Overview of the clonecensorweighting Package for R"
author: "Jihyeon Baek, Hye Won Yang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
    mathjax: null
    pandoc_args: ["--mathml"]
vignette: >
  %\VignetteIndexEntry{An Overview of the clonecensorweighting Package for R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
set.seed(1)
```

```{r setup, include = FALSE}
library(clonecensorweighting)
```

<style>
/* html_vignette's stylesheet sets h1 { margin-top: 0 }, so each numbered
   section starts flush against the preceding text. Give the headings room. */
h1 { margin-top: 2.5em; }
h1.title { margin-top: 0; }   /* keep the document title at the top */
h2 { margin-top: 1.8em; }
h3 { margin-top: 1.4em; }
</style>

# Introduction

`clonecensorweighting` implements the *clone-censor-weight* (CCW) procedure for
emulating a target trial from observational time-to-event data.

Suppose we want to know whether surgery improves survival in lung cancer
patients. Naively comparing those who had surgery against those who did not can
be badly misleading. A central culprit is *immortal time bias*: to be counted in
the "surgery" group, a patient must first survive long enough to receive
surgery, and that guaranteed survival time is then credited to the surgery group
as though it were a benefit of the operation.

CCW is the standard target-trial-emulation remedy for this bias, and its name
traces the three steps it takes:

- **Clone** -- each patient is duplicated into every treatment strategy being
  compared, so that at the start of follow-up every clone is compatible with the
  strategy it is assigned to.
- **Censor** -- a clone is censored the moment the patient's observed history
  stops being consistent with the strategy that clone represents.
- **Weight** -- inverse-probability-of-censoring weights undo the selection bias
  introduced by that artificial censoring.

The package exposes these steps as a small set of composable functions: they
clone a subject-level dataset across strategies, apply the artificial censoring
each strategy implies, expand the result into the person-time format that an
inverse-probability-of-censoring weighted (IPCW) analysis requires, and fit the
weighted per-protocol effect.

The rest of this document is organised as follows. Section 2 reviews target
trial emulation and introduces the data. Section 3 works through the method on
that data, presenting each of the three ideas -- cloning, artificial censoring,
and IPC weighting -- with its rationale and immediately in code, building up the
analysis dataset step by step and then estimating the weighted per-protocol
effect. Section 4 summarises the workflow.

We assume familiarity with the basic ideas of target trial emulation and
inverse-probability weighting. The target-trial framework we emulate is set out
by Hernán and Robins [2016] and Hernán, Wang and Leaf [2022]; marginal structural
models and the use of IPC weighting to adjust for time-varying selection are
developed by Robins, Hernán and Brumback [2000] and applied by Hernán, Brumback
and Robins [2000], with inverse-probability-of-censoring weighting itself
introduced by Robins and Finkelstein [2000]. Hernán [2018] sets out the
cloning-censoring-weighting strategy for treatment-duration questions of the kind
considered here. Maringe et al. [2020] discuss trial emulation specifically as a
remedy for immortal-time bias, and Gaber et al. [2024] give a practical account of
implementing clone-censor-weight with stabilized IPCW, which the implementation
here follows.

# Target trial emulation

## The problem it solves

A randomized controlled trial (RCT) can directly compare "following strategy A
to the end" versus "following strategy B to the end." The difficulty with
observational data is that no such assignment exists. People are not split into
A/B at baseline; over time they begin, switch, or stop treatments according to
their evolving health.

The hardest case is when the strategy is defined by something that happens in the
future, not at baseline. Consider the strategy "receive surgery within 6 months
of diagnosis." At diagnosis (baseline) we cannot know whether a person will go on
to have surgery within 6 months. If we split people at baseline into "surgery"
versus "no surgery," anyone who dies before surgery is automatically classified
into the no-surgery arm, producing *immortal time bias*.

Hernán and Robins [2016] proposed a simple remedy: specify the RCT you would
ideally run (the *target trial*), then emulate that trial using observational
data. A target trial is specified through seven components. Table 1 maps those
components onto this vignette's running example (surgery within 6 months of
diagnosis).

Table: Target trial components mapped onto the running example.

| Target trial component | Definition in this example |
|---|---|
| 1. Eligibility        | Patients alive at diagnosis and not yet operated on |
| 2. Treatment strategies | (A) surgery within 6 months / (B) no surgery within 6 months |
| 3. Assignment         | No assignment in observational data $\rightarrow$ replaced by cloning |
| 4. Follow-up          | From baseline (diagnosis) to death or end of observation |
| 5. Outcome            | Overall survival |
| 6. Causal contrast    | Per-protocol effect (effect of following the strategy to the end) |
| 7. Analysis           | Artificial censoring + IPCW, then weighted survival analysis |

## Data

The `clonecensorweighting` package is concerned with the causal analysis of
time-to-event outcomes. The bundled `lungcancer` dataset contains 200 simulated
lung cancer patients followed for up to one year after diagnosis; 106 received
surgery within six months and 48 died during follow-up.

```{r data}
data(lungcancer)
head(lungcancer)
```

The columns the analysis relies on are:

- `id`: patient identifier
- `surgery`: whether the patient received surgery (`1`) or not (`0`)
- `timetosurgery`: time from baseline to surgery (`NA` if never operated on)
- `death`: whether the patient died (`1`) or not (`0`) -- the outcome event
- `fup_obs`: observed follow-up time

The remaining columns (`sex`, `charlson`, `perf`, `stage`, `emergency`, `age`,
`deprivation`) describe patient characteristics that can be used as covariates
when estimating censoring weights. Below we use `age` and `sex`.

## Reading your own data

The example uses the bundled `lungcancer` data, but the same workflow runs on any
subject-level dataset that provides, at a minimum:

- an identifier column for each patient;
- a binary treatment column, coded `0` / `1`;
- a time-to-treatment column (numeric, `NA` for the untreated);
- a binary outcome column, coded `0` / `1`;
- a numeric follow-up-time column.

The column *names* are up to you: you pass them to the relevant function
arguments (the treatment, outcome and follow-up names to the policy and
censoring helpers; the identifier to `create_final_data()`; any covariates to
`estimate_censoring()`), as we do in Section 3. Any additional columns are
carried along and can serve as covariates.

`read_trial_data()` is a small convenience wrapper for reading such data from a
CSV file into a tibble. It imposes no particular column structure; it simply
reads the file:

```{r read-trial-data, eval = FALSE}
my_data <- read_trial_data("path/to/your-data.csv")
```

# Clone-censor-weighting

We now walk through the method on the `lungcancer` data. Each of the three ideas
that give the method its name -- clone, censor, weight -- is presented with its
rationale and then applied immediately in code, checking the result at each turn.
A final step estimates the effect from the clone-censor-weighted data. We compare
two strategies with a six-month grace period ($\approx 182.62$ days).

The **grace period** is central to this analysis: it is the window during which a
patient is still considered compatible with the "surgery" strategy. A patient who
has surgery within the grace period is consistent with the treated strategy; one
who does not is consistent with the control strategy. Hernán [2018] discusses how
such duration strategies should be specified and emulated.

```{r setup-arms}
arms         <- c("Control", "Surgery")
grace_period <- 182.62
```

## Clone

Each eligible individual is duplicated once per strategy being compared. With two
strategies, one person becomes two clones. At baseline nobody has yet violated
any strategy, so at the moment of duplication both clones share an identical
history. This makes the baseline covariates *exactly* balanced across arms,
removing the source of immortal time bias.

`clone_arms()` carries this out, returning a list with one data frame per arm. At
this point the two are identical copies of the original data; the
strategy-specific logic comes next.

```{r clone}
clones <- clone_arms(lungcancer, arms)
names(clones)
sapply(clones, nrow)
```

## Censor

Each clone is artificially censored the first moment it deviates from the
strategy it represents.

- The "surgery within 6 months" clone: if 6 months pass without surgery, the
  clone has now violated its strategy and is censored at the 6-month mark. If
  surgery occurs in time, the clone remains in the risk set.
- The "no surgery within 6 months" clone: if surgery occurs within 6 months, the
  clone is censored at the time of surgery.

In other words, each clone stays under follow-up only while it remains consistent
with its strategy.

### The intuition, traced by hand

Before applying the package functions, it helps to trace the censoring by hand on
a small dataset. The bundled `patients13` dataset is exactly that: the 13 patient
records of Figure 2 in Maringe et al. [2020], chosen to cover every censoring
pattern that can arise in registry data. Times are in days, on the same scale as
`lungcancer`.

```{r patients13-data}
data(patients13)
patients13
```

Here `time_to_surgery` is the time to surgery (`NA` if the patient never had
surgery), `followup` is the observed follow-up time, and `death` is the event
indicator ($1$ = death).

Using the same six-month `grace_period` as above, we work out when each clone is
censored. The one subtlety is that a clone can only *deviate* from a strategy if
it is still under observation at the moment the deviation would occur: a patient
whose follow-up ends before the grace period closes never gets the chance to
violate the "surgery within 6 months" strategy, so their observed outcome still
counts.

```{r patients13-hand-trace}
## --- Surgery-arm clone: surgery must occur within the grace period ---
##   * surgery in time                -> adherent, follow to end of observation
##   * still unoperated when the
##     grace period closes            -> censored there
##   * follow-up ends before that     -> no deviation, observed outcome stands
surg_clone <- within(patients13, {
  adherent <- !is.na(time_to_surgery) & time_to_surgery <= grace_period
  deviates <- !adherent & followup > grace_period
  fu_time  <- ifelse(deviates, grace_period, followup)
  status   <- ifelse(deviates, 0, death)   # censored -> status = 0
  arm      <- "Surgery"
})

## --- Control-arm clone: surgery must NOT occur within the grace period ---
##   * surgery within the grace period -> censored at the time of surgery
##   * otherwise                       -> adherent, follow to end of observation
ctrl_clone <- within(patients13, {
  deviates <- !is.na(time_to_surgery) & time_to_surgery <= grace_period
  fu_time  <- ifelse(deviates, time_to_surgery, followup)
  status   <- ifelse(deviates, 0, death)
  arm      <- "Control"
})

cols <- c("id", "arm", "fu_time", "status")
rbind(surg_clone[cols], ctrl_clone[cols])
```

Three rows carry the whole idea. Patient A has surgery on day 61: adherent in the
surgery arm and followed to death on day 300, but censored on day 61 in the
control arm. Patient F is the mirror image -- surgery on day 220, after the grace
period has closed -- so the surgery-arm clone is censored at day 182.62 while the
control-arm clone remains adherent and contributes its death on day 320. Patient
K never has surgery and dies on day 40, before the grace period closes: neither
strategy has been violated, so that death counts in *both* arms. This table is
the core intuition of CCW.

### Deriving the grace-period rules

The package derives exactly this censoring from the grace-period rules, on the
real data. Two helper functions describe how each arm's emulated outcome,
follow-up, and censoring should be derived. Note that they do not touch the data
yet: they return the *rules* (as expressions), one nested list entry per arm,
which are evaluated in the next step.

`create_policy_A()` produces the rules for the emulated outcome and follow-up
under the grace-period policy, and `create_censoring_logics_A()` produces the
rules for the censoring indicator and the time at which that indicator can first
be determined. Here we name the resulting columns explicitly (`outcome`, `fup`,
`censoring`, `fup_uncensored`); if these arguments are omitted, the dot-prefixed
names `.outcome`, `.fup`, `.censoring` and `.fup_uncensored` are used instead.

`apply_logics()` evaluates a set of rules against each arm's data, so we call it
once per rule set.

```{r censor-apply}
policies <- create_policy_A(
  arms,
  treatment         = "surgery",
  time_to_treatment = "timetosurgery",
  grace_period      = grace_period,
  outcome           = "death",
  followup          = "fup_obs",
  clone_outcome     = "outcome",
  clone_followup    = "fup"
)
clones_policy <- apply_logics(clones, policies)

censoring_logics <- create_censoring_logics_A(
  arms,
  treatment                 = "surgery",
  time_to_treatment         = "timetosurgery",
  grace_period              = grace_period,
  followup                  = "fup_obs",
  clone_censoring           = "censoring",
  clone_uncensored_followup = "fup_uncensored"
)
clones_censored <- apply_logics(clones_policy, censoring_logics)

setdiff(names(clones_censored$Surgery), names(lungcancer))
```

Each arm's data frame now carries these four emulated variables alongside the
original columns.

### Expanding to person-time

Finally, `create_final_data()` expands each clone into a counting-process
("long") table -- one row per patient-time interval, bounded by `Tstart` and
`Tstop` -- the format the weighting step requires. Internally it finds every
event time, splits each patient's follow-up at those times, and combines the
outcome and censoring information into one table per arm. Because each patient
contributes several intervals, each arm has many more rows than the original 200
patients.

```{r censor-final}
clones_final <- create_final_data(
  clones_censored,
  clone_followup  = "fup",
  clone_outcome   = "outcome",
  clone_censoring = "censoring",
  col_ids         = "id"
)
head(clones_final$Surgery)
sapply(clones_final, nrow)
```

## Weight

The catch is that this artificial censoring is *not random*. Whether a clone
deviates (and is censored) is related to its evolving health. If, say, sicker
patients are more or less likely to have surgery, the censoring introduces
selection bias, which *inverse-probability-of-censoring weighting* (IPCW)
corrects [Robins and Finkelstein 2000].

Some notation first. Index individuals by $i$ and discrete follow-up times by
$k = 1, 2, \dots$; let $L_0$ be the baseline covariates and
$\bar L_k = (L_0, \dots, L_k)$ the covariate history through time $k$. For a clone
in a given arm, let $C_k = 1$ mean it is artificially censored at time $k$, and
$\bar C_{k-1} = 0$ that it was still uncensored up to time $k-1$. The probability
of being censored in an interval is modelled by pooled logistic regression on the
person-time data -- the discrete-time hazard approach of Efron [1988], whose close
correspondence with time-dependent Cox regression is established by D'Agostino et
al. [1990] --

$$
\operatorname{logit}\Pr\!\big(C_k = 1 \mid \bar C_{k-1}=0,\; \bar L_k,\; \text{arm}\big)
\;=\; f(t_k) + \beta^\top L_k,
$$

with $t_k$ the interval start time. By default, $f(t_k)$ is linear. Setting
`time_spline_df = 3` uses a natural cubic spline with 3 degrees of freedom,
allowing the baseline log-odds of censoring to vary nonlinearly over follow-up.
Spline flexibility should be supported by enough censoring events; sparse or
structurally concentrated censoring can otherwise cause separation. IPCW then
multiplies each clone by the inverse probability of "remaining uncensored so
far," reallocating the contribution of censored clones to similar clones that
remain. The **unstabilized** weight is

$$
W_i(k) \;=\;
\frac{1}{\displaystyle\prod_{j=1}^{k}
  \Pr\!\big(C_j = 0 \,\mid\, \bar C_{j-1}=0,\; \bar L_j,\; \text{arm}\big)},
$$

and the **stabilized** weight replaces the constant numerator by a baseline-only
model -- the same regression, but with $L_0$ in place of $\bar L_j$:

$$
SW_i(k) \;=\;
\prod_{j=1}^{k}
\frac{\Pr\!\big(C_j = 0 \,\mid\, \bar C_{j-1}=0,\; \bar L_0,\; \text{arm}\big)}
     {\Pr\!\big(C_j = 0 \,\mid\, \bar C_{j-1}=0,\; \bar L_j,\; \text{arm}\big)}.
$$

The *denominator* conditions on the full time-varying history $\bar L_j$ and
models the actual censoring mechanism; the *numerator* conditions on baseline
covariates $\bar L_0$ only, reducing the variance of the weights without changing
the causal estimand; this stabilization is due to Robins, Hernán and Brumback
[2000]. Taking the cumulative product across time gives $W_i(k)$ or $SW_i(k)$.

In the package, `estimate_censoring()` fits these probabilities and adds the
uncensoring probability `P_uncens` to each row; `weight_cases()` forms the weight
in the column `weight_Cox`. Here `age` and `sex` are fixed at baseline and there are no time-varying covariates, so a stabilized weight with a baseline-only numerator would collapse to 1 at every interval and remove the adjustment entirely. We therefore use the unstabilized weight, whose denominator still adjusts for the `age`/`sex` dependence of censoring.

```{r weight}
clones_estimated <- estimate_censoring(
  clones_final,
  predictors = c("age", "sex"),
  method = "pooled_logit"
)
clones_weighted <- weight_cases(clones_estimated)
```

It is good practice to inspect the weight distribution: extremely large weights
inflate variance and can signal near-violations of positivity. Cole and Hernán
[2008] give practical guidance on constructing and checking such weights.

```{r weight-diagnostics, fig.width = 6, fig.height = 4, fig.cap = "Distribution of IPC weights."}
weights_all <- unlist(
  lapply(clones_weighted, function(x) x[["weight_Cox"]]),
  use.names = FALSE
)
summary(weights_all)
hist(weights_all, breaks = 40, col = "grey80", border = "white",
     main = NULL, xlab = "IPC weight")
```

## Estimating the effect

With clone-censor-weighted data in hand, `emul_estimate()` fits the weighted Cox
model

$$
\lambda\!\big(t \mid \text{arm}, L\big)
\;=\; \lambda_0(t)\,\exp\!\big\{\gamma\,\mathbb{1}(\text{arm}=\text{treated}) + \delta^\top L\big\},
$$

weighting each person-time record by its IPC weight. The per-protocol hazard
ratio is $\mathrm{HR} = e^{\gamma}$ -- the causal contrast specified in Section 2.

```{r cox}
cox_fit <- emul_estimate(
  clones_weighted,
  method     = "Cox",
  weights    = "weight_Cox",
  predictors = c("age", "sex")
)
summary(cox_fit)$conf.int
```

Because the weights make the effective sample differ from the raw data, the
model-based standard error is unreliable. `emul_estimate_bootstrap()` resamples
patients from the original subject-level data $B$ times. Within every resample,
it repeats cloning, artificial censoring, person-time expansion, censoring-model
estimation, weighting, and outcome-model estimation before taking the
percentile interval [Efron and Tibshirani 1993] of the resulting hazard ratios,

$$
\big(\widehat{\mathrm{HR}}^{*}_{(\alpha/2)},\ \widehat{\mathrm{HR}}^{*}_{(1-\alpha/2)}\big),
$$

at confidence level $1-\alpha$. The number of resamples $B$ is kept small here
for speed.

```{r bootstrap}
boot <- emul_estimate_bootstrap(
  lungcancer,
  arms                   = arms,
  id                     = "id",
  treatment              = "surgery",
  time_to_treatment      = "timetosurgery",
  grace_period           = grace_period,
  outcome                = "death",
  followup               = "fup_obs",
  censoring_predictors   = c("age", "sex"),
  method                 = "Cox",
  predictors             = c("age", "sex"),
  n_bootstrap            = 10,
  seed                   = 1
)
c(
  HR = boot$estimate,
  HR_lower = boot$ci_lower,
  HR_upper = boot$ci_upper
)
```

For a visual summary, `method = "KM"` returns weighted Kaplan-Meier curves for
the two strategies.

```{r km, fig.width = 6, fig.height = 4, fig.cap = "Weighted survival curves by strategy."}
km_fit <- emul_estimate(clones_weighted, method = "KM", weights = "weight_Cox")
plot(km_fit, col = c("#1b9e77", "#d95f02"), lwd = 2,
     xlab = "Days since time zero", ylab = "Survival probability")
legend("bottomleft", legend = arms, col = c("#1b9e77", "#d95f02"),
       lwd = 2, bty = "n")
```

# Summary

Working through the surgery-in-lung-cancer question, this vignette:

- motivated clone-censor-weighting with the problem of immortal time bias, and
  specified the target trial being emulated (Section 2);
- cloned the subject-level data across a control and a surgery arm with
  `clone_arms()`;
- derived the grace-period policy and artificial-censoring rules with
  `create_policy_A()` and `create_censoring_logics_A()`, and evaluated them with
  `apply_logics()`;
- expanded the result into a person-time analysis dataset with
  `create_final_data()`;
- estimated censoring probabilities with `estimate_censoring()` and formed IPC
  weights with `weight_cases()`, inspecting their distribution;
- and estimated the weighted per-protocol effect with `emul_estimate()`, with a
  percentile confidence interval from `emul_estimate_bootstrap()` and weighted
  Kaplan-Meier curves for a visual summary.

The same sequence applies to any dataset with the five columns listed in
Section 2: an identifier, a binary treatment, a time to treatment, a binary
outcome, and a follow-up time.

# References

Cole, S. R., & Hernán, M. A. (2008). Constructing inverse probability weights for
marginal structural models. *American Journal of Epidemiology*, 168(6), 656-664.

D'Agostino, R. B., Lee, M.-L., Belanger, A. J., Cupples, L. A., Anderson, K., &
Kannel, W. B. (1990). Relation of pooled logistic regression to time dependent Cox
regression analysis: the Framingham Heart Study. *Statistics in Medicine*, 9(12),
1501-1515.

Efron, B. (1988). Logistic regression, survival analysis, and the Kaplan-Meier
curve. *Journal of the American Statistical Association*, 83(402), 414-425.

Efron, B., & Tibshirani, R. J. (1993). *An Introduction to the Bootstrap*.
Chapman & Hall, New York.

Gaber, C. E., Ghazarian, A. A., Strassle, P. D., Ribeiro, T. B., Salas, M., &
Maringe, C. (2024). De-mystifying the clone-censor-weight method for causal
research using observational data: a primer for cancer researchers.
*Cancer Medicine*, 13(23), e70461.

Hernán, M. A. (2018). How to estimate the effect of treatment duration on survival
outcomes using observational data. *BMJ*, 360, k182.

Hernán, M. A., Brumback, B., & Robins, J. M. (2000). Marginal structural models
to estimate the causal effect of zidovudine on the survival of HIV-positive men.
*Epidemiology*, 11(5), 561-570.

Hernán, M. A., & Robins, J. M. (2016). Using big data to emulate a target trial
when a randomized trial is not available. *American Journal of Epidemiology*,
183(8), 758-764.

Hernán, M. A., Wang, W., & Leaf, D. E. (2022). Target trial emulation: a
framework for causal inference from observational data. *JAMA*, 328(24),
2446-2447.

Maringe, C., Benitez Majano, S., Exarchakou, A., et al. (2020). Reflection on
modern methods: trial emulation in the presence of immortal-time bias.
*International Journal of Epidemiology*, 49(5), 1719-1729.

Robins, J. M., & Finkelstein, D. M. (2000). Correcting for noncompliance and
dependent censoring in an AIDS clinical trial with inverse probability of
censoring weighted (IPCW) log-rank tests. *Biometrics*, 56(3), 779-788.

Robins, J. M., Hernán, M. A., & Brumback, B. (2000). Marginal structural models
and causal inference in epidemiology. *Epidemiology*, 11(5), 550-560.
